---
title: 'DS202: Lab 4'
author: "MyTien Kien; kmytien"
date: "3/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
defense <-  read_excel('C:\\Users\\kmyti\\OneDrive\\Desktop\\ds202_excel\\cyclonesFootball2020.xlsx', sheet='Defensive')
offense <-  read_excel('C:\\Users\\kmyti\\OneDrive\\Desktop\\ds202_excel\\cyclonesFootball2020.xlsx', sheet='Offensive')
bio <-  read_excel('C:\\Users\\kmyti\\OneDrive\\Desktop\\ds202_excel\\cyclonesFootball2020.xlsx', sheet='Biography')
```

## Part 1: Cleaning
```{r message=FALSE}
defClean <- defense %>%
      mutate(Name = factor(Name)) %>%
      mutate(Opponent_Opponent = factor(Opponent_Opponent)) %>%
      mutate(across(3:11, as.numeric))

offClean <- offense %>%
      mutate(Name = factor(Name)) %>%
      mutate(Opponent_Opponent = factor(Opponent_Opponent)) %>%
      mutate(across(3:12, as.numeric))

bio <- bio %>%
      mutate(Name = factor(Name)) %>%
      mutate(Weight = as.numeric(Weight))

bioClean <- bio %>%
      separate(Height, into = c("Ft", "In"), convert = TRUE) %>%
      mutate(Height = (Ft * 12) + In) %>%
      select(-c(Ft:In))

str(defClean)
str(offClean)
str(bioClean)
```

//say something

## Part 2: Tidying

**1.** Reformat the offClean data frame into a tidy format using pivot_longer. The type of statistic (Receiving_REC for the number of receivings, Receiving_YDS for receiving yards) is added as a new key column named stat.
```{r message=FALSE}
p_one <- offClean %>%
      pivot_longer(c(3:4), names_to = "Stat", values_to = "Value")
```

**2.** Show on the distributions of the offensive statistics. The distribution of each offensive statistic should be shown in a separate facet. Explain the patterns you see.
```{r}
p_two <- p_one %>% 
      ggplot(aes(x = Value)) + geom_histogram() + facet_wrap(~Stat, scales = 'free_y')

p_two + ggtitle('Distributions of Offense Stats') + ylab("Count of Receivings/Yards") + xlab("Value")
```

//say something

**3.** Did ISU have better offense against Oregon or Oklahoma? Answer this question by creating a scatterplot with the x- and y-axes being the receiving yards of the ISU players against these two opponents. A large number of receiving yards indicates good offense.
```{r}

```

//say something

**4.** Separate the Hometown column into two columns, namely, the city and the state where the player comes from. Print the first few rows of these two columns. (Hint: look at the sep= argument of separate)
```{r}
p_four <- bioClean %>%
      separate(Hometown, into = c('City', 'State'), sep = ',')

head(p_four)
```

**5.** How many players are there from each state?
```{r}
p_five <- p_four %>%
      group_by(State) %>%
      summarise(Players = n())

p_five
```

## Part 3: Joining Data Frames
**1.** Find which states do the ISU players (both defensive and offensive) come from. Create an appropriate numerical or graphical summary.
```{r}

```

**2.** How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?
```{r}

```

**3.** Which ISU player(s) made the biggest progress over the last year? Compare the player-level statistics (i.e. averages and totals for each player) using both the 2020 and the 2019 data here. Choose for yourself the player statistics to investigate.
```{r}

```