---
title: 'DS202: Lab 4'
author: "MyTien Kien; kmytien"
date: "3/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
defense <-  read_excel('C:\\Users\\kmyti\\OneDrive\\Desktop\\ds202_excel\\cyclonesFootball2020.xlsx', sheet='Defensive')
offense <-  read_excel('C:\\Users\\kmyti\\OneDrive\\Desktop\\ds202_excel\\cyclonesFootball2020.xlsx', sheet='Offensive')
bio <-  read_excel('C:\\Users\\kmyti\\OneDrive\\Desktop\\ds202_excel\\cyclonesFootball2020.xlsx', sheet='Biography')
```

## Part 1: Cleaning
```{r message=FALSE}
#Factored and converted necessary columns to numeric for Defense and Offense
defClean <- defense %>%
      mutate(Name = factor(Name)) %>%
      mutate(Opponent_Opponent = factor(Opponent_Opponent)) %>%
      mutate(across(3:11, as.numeric))

offClean <- offense %>%
      mutate(Name = factor(Name)) %>%
      mutate(Opponent_Opponent = factor(Opponent_Opponent)) %>%
      mutate(across(3:12, as.numeric))

#For Bio, I converted height into inches, as well as factored Name and turned Weight numeric
bio <- bio %>%
      mutate(Name = factor(Name)) %>%
      mutate(Weight = as.numeric(Weight))

bioClean <- bio %>%
      separate(Height, into = c("Ft", "In"), convert = TRUE) %>%
      mutate(Height = (Ft * 12) + In) %>%
      select(-c(Ft:In))

#Printing Structures
str(defClean)
str(offClean)
str(bioClean)
```

For defClean & offClean, I converted the Name & Opponent so that it is factored, and everything else is numeric. For bioClean, I factored the Name as well as turned Weight numerical. For The Height in bioClean, I chose to convert the height into inches.


## Part 2: Tidying
**1.** Reformat the offClean data frame into a tidy format using pivot_longer. The type of statistic (Receiving_REC for the number of receivings, Receiving_YDS for receiving yards) is added as a new key column named stat.
```{r message=FALSE}
p_one <- offClean %>%
      pivot_longer(c(3:12), names_to = "Stat", values_to = "Value")
```


**2.** Show on the distributions of the offensive statistics. The distribution of each offensive statistic should be shown in a separate facet. Explain the patterns you see.
```{r message=FALSE}
p_two <- p_one %>% 
      ggplot(aes(x = Value)) + geom_histogram(bins = 10) + facet_wrap(~Stat, scales = 'free_y')

p_two + ggtitle('Distributions of Offense Stats') + ylab("Count of Specific Offense Stat") + xlab("Value")
```

//say something


**3.** Did ISU have better offense against Oregon or Oklahoma? Answer this question by creating a scatter plot with the x- and y-axes being the receiving yards of the ISU players against these two opponents. A large number of receiving yards indicates good offense.
```{r message=FALSE}
#Not sure if I did this correctly
p_three <- p_one %>%
         select(Opponent_Opponent, Stat, Value) %>%
         filter(Opponent_Opponent == 'Oklahoma' | Opponent_Opponent == 'Oregon', Stat == 'Receiving_YDS') %>%
         ggplot(aes(x = Value, y = Opponent_Opponent, color = Opponent_Opponent)) + geom_point(position = 'jitter')

p_three
```

From this graph, it looks like Iowa State's Offense did a lot better against Oklahoma than Oregon.


**4.** Separate the Hometown column into two columns, namely, the city and the state where the player comes from. Print the first few rows of these two columns. (Hint: look at the sep= argument of separate)
```{r}
p_four <- bioClean %>%
      separate(Hometown, into = c('City', 'State'), sep = ',')

head(select(p_four, Name, City, State))
```


**5.** How many players are there from each state?
```{r}
p_five <- p_four %>%
      group_by(State) %>%
      summarise(Players = n())

p_five
```

From this, all players on the ISU Football team are from a total of 21 different states. The most being from Iowa.


## Part 3: Joining Data Frames

**1.** Find which states do the ISU players (both defensive and offensive) come from. Create an appropriate numerical or graphical summary.
```{r}
#Offense
offState <- offClean %>%
      distinct(Name) %>%
      left_join(p_four, by = c("Name" = "Name"))

offState %>%
      group_by(State) %>% 
      summarize(Players = n()) %>%
      ggplot(aes(x = State, weight = Players)) + geom_bar() + coord_flip() +
      ggtitle("Offense Players and States") + ylab("Number of Offense Players")
```

```{r}
#Defense
defState <- defClean %>%
      distinct(Name) %>%
      left_join(p_four, by = c("Name" = "Name"))

defState %>%
      group_by(State) %>% 
      summarize(Players = n()) %>%
      ggplot(aes(x = State, weight = Players)) + geom_bar() + coord_flip() +
      ggtitle("Defense Players and States") + ylab("Number of Defense Players")
```


**2.** How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?
```{r}

```


**3.** Which ISU player(s) made the biggest progress over the last year? Compare the player-level statistics (i.e. averages and totals for each player) using both the 2020 and the 2019 data here. Choose for yourself the player statistics to investigate.
```{r}

```