---
title: 'DS202: Lab 4'
author: "MyTien Kien; kmytien, Tanner Dunn; tcdunn"
date: "3/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(readxl)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
defense <-  read_excel('C:\\Spring2021\\ds202\\lab4\\cyclonesFootball2020.xlsx', sheet='Defensive')
offense <-  read_excel('C:\\Spring2021\\ds202\\lab4\\cyclonesFootball2020.xlsx', sheet='Offensive')
offense19 <-  read_excel('C:\\Spring2021\\ds202\\lab4\\cyclonesFootball2019.xlsx', sheet='Offensive')
bio <-  read_excel('C:\\Spring2021\\ds202\\lab4\\cyclonesFootball2020.xlsx', sheet='Biography')
```

## Part 1: Cleaning
```{r message=FALSE}
#Factored and converted necessary columns to numeric for Defense and Offense
defClean <- defense %>%
      mutate(Name = factor(Name)) %>%
      mutate(Opponent_Opponent = factor(Opponent_Opponent)) %>%
      mutate(across(3:11, as.numeric))

offClean <- offense %>%
      mutate(Name = factor(Name)) %>%
      mutate(Opponent_Opponent = factor(Opponent_Opponent)) %>%
      mutate(across(3:12, as.numeric))

#For Bio, I converted height into inches, as well as factored Name and turned Weight numeric
bio <- bio %>%
      mutate(Name = factor(Name)) %>%
      mutate(Weight = as.numeric(Weight))

bioClean <- bio %>%
      separate(Height, into = c("Ft", "In"), convert = TRUE) %>%
      mutate(Height = (Ft * 12) + In) %>%
      select(-c(Ft:In))

#Printing Structures
str(defClean)
str(offClean)
str(bioClean)
```

For defClean & offClean, I converted the Name & Opponent so that it is factored, and everything else is numeric. For bioClean, I factored the Name as well as turned Weight numerical. For The Height in bioClean, I chose to convert the height into inches.


## Part 2: Tidying
**1.** Reformat the offClean data frame into a tidy format using pivot_longer. The type of statistic (Receiving_REC for the number of receiving, Receiving_YDS for receiving yards) is added as a new key column named stat.
```{r message=FALSE}
p_one <- offClean %>%
      pivot_longer(c(3:12), names_to = "Stat", values_to = "Value")
```


**2.** Show on the distributions of the offensive statistics. The distribution of each offensive statistic should be shown in a separate facet. Explain the patterns you see.
```{r message=FALSE}
p_two <- p_one %>% 
      ggplot(aes(x = Value)) + geom_histogram(bins = 30) + facet_wrap(~Stat, scales = 'free_y')

p_two + ggtitle('Distributions of Offense Stats') + ylab("Count of Specific Offense Stat") + xlab("Frequency")
```

It looks like there are may variables in the Offense Statistics that consists of values below 10, such as Passing_INT, Passing_TD, Receiving_TD and so on. From most of these graphs, it looks like as we go up the x-value for a any specific offense stat/variable, the less likely we'll see someone achieve that number in that specific stat.


**3.** Did ISU have better offense against Oregon or Oklahoma? Answer this question by creating a scatter plot with the x- and y-axes being the receiving yards of the ISU players against these two opponents. A large number of receiving yards indicates good offense.
```{r message=FALSE}
#Not sure if I did this correctly - possibly use pivot_longer?
p_three <- p_one %>%
         select(Opponent_Opponent, Stat, Value) %>%
         filter(Opponent_Opponent == 'Oklahoma' | Opponent_Opponent == 'Oregon', Stat == 'Receiving_YDS') %>%
         ggplot(aes(x = Value, y = Opponent_Opponent, color = Opponent_Opponent)) + geom_point(position = 'jitter') +
         ggtitle("Receiving Yards vs Oklahoma/Oregon") + ylab("Opponent")

p_three
```

From this graph, it looks like Iowa State's Offense did a lot better against Oklahoma than Oregon. There are some points against Oklahoma that are much higher than Oregon's. The Receiving_Yards go up to 114 for Oklahoma compared for Oregon's highest, ~55 yards.


**4.** Separate the Hometown column into two columns, namely, the city and the state where the player comes from. Print the first few rows of these two columns. (Hint: look at the sep= argument of separate)
```{r}
p_four <- bioClean %>%
      separate(Hometown, into = c('City', 'State'), sep = ',')

head(select(p_four, Name, City, State))
```


**5.** How many players are there from each state?
```{r}
p_five <- p_four %>%
      group_by(State) %>%
      summarise(Players = n())

p_five
```

From this, all players on the ISU Football team are from a total of 21 different states. The most being from Iowa.


## Part 3: Joining Data Frames
**1.** Find which states do the ISU players (both defensive and offensive) come from. Create an appropriate numerical or graphical summary.
```{r}
#Offense - 16 Players in Total
offState <- offClean %>%
      distinct(Name) %>%
      left_join(p_four, by = c("Name" = "Name"))

offState %>%
      group_by(State) %>% 
      summarize(Players = n()) %>%
      ggplot(aes(x = State, weight = Players)) + geom_bar() + coord_flip() +
      ggtitle("Offense Players and States") + ylab("Number of Offense Players")
```

```{r}
#Defense - 37 Players in Total
defState <- defClean %>%
      distinct(Name) %>%
      left_join(p_four, by = c("Name" = "Name"))

defState %>%
      group_by(State) %>% 
      summarize(Players = n()) %>%
      ggplot(aes(x = State, weight = Players)) + geom_bar() + coord_flip() +
      ggtitle("Defense Players and States") + ylab("Number of Defense Players")
```


**2.** How does the performance of quarterback Brock Purdy relate to the performance of the ISU team?
```{r}
both <- left_join(offClean, defClean)

brock <- both %>% 
   filter(Name == 'Purdy, Brock') %>%
   summarise(Pass_mean = mean(Passing_YDS), Rush_mean = mean(Rushing_YDS))

team <- both %>%
   filter(Name != 'Purdy, Brock') %>%
   group_by(Name) %>%
   summarise(Rec_mean = mean(Receiving_YDS, rm.na = TRUE), Rush_mean = mean(Rushing_YDS, rm.na = TRUE))

brock
team

```
Looking at Brock Purdy's performance in relation to his teammates you can see he averages around 238 passing Yards a game and the player who
averages the largest portion of those yards is Xavier Hutchinson with 69.1. Also looking at his rushing yards Breece Hall is the only one on average
out rushing him which makes sense as he is the star running back. 


**3.** Which ISU player(s) made the biggest progress over the last year? Compare the player-level statistics (i.e. averages and totals for each player) using both the 2020 and the 2019 data here. Choose for yourself the player statistics to investigate.
```{r}
offense19 <-  read_excel('C:\\Spring2021\\ds202\\lab4\\cyclonesFootball2019.xlsx', sheet='Offensive')
defense19 <- read_excel('C:\\Spring2021\\ds202\\lab4\\cyclonesFootball2019.xlsx', sheet='Defensive')

defClean19 <- defense19 %>%
      mutate(Name = factor(Name)) %>%
      mutate(Opponent_Opponent = factor(Opponent_Opponent)) %>%
      mutate(across(3:11, as.numeric))

offClean19 <- offense19 %>%
      mutate(Name = factor(Name)) %>%
      mutate(Opponent_Opponent = factor(Opponent_Opponent)) %>%
      mutate(across(3:12, as.numeric))

defense <-  read_excel('C:\\Spring2021\\ds202\\lab4\\cyclonesFootball2020.xlsx', sheet='Defensive')
offense <-  read_excel('C:\\Spring2021\\ds202\\lab4\\cyclonesFootball2020.xlsx', sheet='Offensive')


defClean <- defense %>%
      mutate(Name = factor(Name)) %>%
      mutate(Opponent_Opponent = factor(Opponent_Opponent)) %>%
      mutate(across(3:11, as.numeric))

offClean <- offense %>%
      mutate(Name = factor(Name)) %>%
      mutate(Opponent_Opponent = factor(Opponent_Opponent)) %>%
      mutate(across(3:12, as.numeric))

both20 <- left_join(defClean, offClean)
both19 <- left_join(defClean19, offClean19)

thisYear <- both20 %>%
   group_by(Name) %>%
   summarise(mean_Rec=mean(Receiving_YDS, rm.na = TRUE), mean_Rush = mean(Rushing_YDS, rm.na = TRUE), mean_Tackles = mean(Tackles_Solo, rm.na= TRUE),
             tot_Rec=sum(Receiving_YDS, rm.na = TRUE), tot_Rush = sum(Rushing_YDS, rm.na = TRUE), tot_Tackles = sum(Tackles_Solo, rm.na= TRUE))

lastYear <- both19 %>%
   group_by(Name) %>%
   summarise(mean_Rec=mean(Receiving_YDS, rm.na = TRUE), mean_Rush = mean(Rushing_YDS, rm.na = TRUE), mean_Tackles = mean(Tackles_Solo, rm.na= TRUE),
             tot_Rec=sum(Receiving_YDS, rm.na = TRUE), tot_Rush = sum(Rushing_YDS, rm.na = TRUE), tot_Tackles = sum(Tackles_Solo, rm.na= TRUE))

head(thisYear, 20)

head(lastYear, 20)

```

Looking at these two tables there are a handful of players who made great improvement.
Akers, Landen tripled his total receiving yards on the offensive side
Jake, Hummel also had great improvement in his total tackels on the defensive side
Lastly Breece Hall improved greatly in all categories. 